---
title: "PS5-DLJ949"
author: "Dylan Johnson"
date: "2023-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
```

# 1 Load and check data (5pt)
You first task is to do a very simple data check: 

1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating). \

2. (2pt) Load data. How many rows/columns do we have?  \
```{r}
population_data <- read.delim("gapminder.csv")
nrow(population_data)
ncol(population_data)
```

3. (2pt) Print a small sample of data. Does it look OK? 
```{r}
sample_n(population_data, 2)
```
The data printed a little strangely, but that is likely due to how many columns there were.

---------------------------------------------------------------------


# 2 Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.
```{r}
population_data %>% 
  group_by(name) %>% 
  filter(!is.na(name)) %>% 
  summarise(n = n_distinct(name)) %>% 
  arrange(desc(n))
```

There are 250 unique country names.

```{r}
population_data %>% 
  group_by(iso2) %>% 
  filter(!is.na(name)) %>% 
  summarise(n = n_distinct(name)) %>% 
  arrange(desc(n))
```


There are 249 unique country iso2 codes, with two countries having
an iso-2 code of blank.

```{r}
population_data %>% 
  group_by(iso3) %>% 
  filter(!is.na(name)) %>% 
  summarise(n = n_distinct(name)) %>% 
  arrange(desc(n))
```
There are 253 unique iso3 codes.

2. If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
    
    ```{r}
population_data %>% 
  group_by(name) %>% 
  filter(iso2=="" & !is.na(name) & name!="") %>% 
  summarise(n = n_distinct(name))
```
    
  There are two iso-2 codes that are blank. One is from Namibia, and the other is from an undisclosed country.
    
    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
    
```{r}
population_data %>% 
  filter(!is.na(iso3) & name=="") %>% 
  group_by(name) %>% 
  distinct(iso3)
```
    

  
3. (2pt) What is the minimum and maximum year in these data?
```{r}
min(population_data$time, na.rm = TRUE)
max(population_data$time, na.rm = TRUE)
```

---------------------------------------------------------------------

# 3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data? \

```{r}
population_data %>% 
  group_by(time) %>% 
  filter(is.na(co2), is.na(co2_PC)) %>% 
  summarise(num_co2 = length(is.na(co2)),
  num_co2_pc = length(is.na(co2_PC))) %>% 
  arrange(rank(desc(num_co2)))

```
  
2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see. \

```{r}
population_data %>% 
  filter(name == c("United States of America", "China", "India", "Poland", "United Kingdom of Great Britain and Northern Ireland")) %>% 
  ggplot(aes(time, co2, col=factor(name))) +
  geom_line() +
  scale_x_continuous(limits = c(min(population_data$time), max(population_data$time))) +
  scale_y_continuous(limits = c(min(population_data$co2), max(population_data$co2)))
```

We see that the United States during the duration of the Cold War was dominant in its CO2 emissions, but rapidly has been overtaken by the industrializing and new energy production center of the world, China. China has skyrocketed in its harmful CO2 emissions since its turn to communism, but India has also been rapidly increasing its CO2 emissions since the turn of the 21st century. Poland and the UK, however, have stayed relatively low for their entire duration of graphing Co2 emissions.


3. (5pt) Now let’s analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of thesame countries. What does this figure suggest? \

```{r}
population_data %>% 
  filter(name == c("United States of America", "China", "India", "Poland", "United Kingdom of Great Britain and Northern Ireland")) %>% 
  ggplot(aes(time, co2_PC, col=factor(name))) +
  geom_line() +
  scale_x_continuous(limits = c(min(population_data$time), max(population_data$time))) +
  scale_y_continuous(limits = c(min(population_data$co2_PC), max(population_data$co2_PC)))
```

Contrary to the previous graph, the graph of CO2 emissions per capita paints a different picture than simply gross CO2 emissions. China and India's monstrous populations make it so that the average Chinese or Indian citizen does not produce much CO2, especially relative to those found in Europe and especially in the US. Poland surprisingly overtakes the UK during the 80s in C)2 emissions per capita but has been decreasing, while the UK has been consistently decreasing CO2 emissions per capita and is now even below China. The US, however, is a massive outlier and the average American easily produces much more CO2 than many of these nations combined when looking at per capita emissions.

4. (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see. \
Note: just compute averages over countries and ignore the fact that countries are of different size. \
Hint: Americas 2016 should be 4.80. \

```{r}
population_data %>% 
  select(region, co2_PC, time) %>% 
  filter(!is.na(region), !is.na(co2_PC), !region=="", time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  summarise(avg_col2_PC = mean(co2_PC, na.rm = TRUE)) %>% 
  ggplot(aes(region, avg_col2_PC, fill=factor(time))) +
  geom_col(position="dodge") +
  labs(title = "Average CO2 Emissions per capita by Continent",
       x="Continent",
       y="Average CO2 per capita (Metric Ton per Capita)",
       fill="Time (Year)")
```


5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. \
Hint: it should look something along these lines(**Please refer to the spec for the graph**) \

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent). \

  We are missing data for 2019 for most countries, but for 2016 the three largest CO2 emitters per capita continents would be Europe, Asia, and the Americas in that order. The three smallest CO2 emitters (least to greatest) would be Africa, Oceania, and then the Americas. This data, however, filtered out the nameless continent with the large CO2 emissions, so had that counted, the top three would be the nameless contintent, Europe, and Asia and the three smallest would be Africa, Oceania, and the Americas in 2016.

---------------------------------------------------------------------

# 4 GDP per capita (50pt)
Let’s look at GDP per capita (*GDP_PC*)

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. \
Comment what do you see there. \

```{r}
gdp_v_le_1960 <- population_data %>% 
  select(name, region, GDP_PC, lifeExpectancy, time, totalPopulation) %>% 
  filter(!is.na(name), !is.na(GDP_PC), !is.na(totalPopulation), !name=="", 
         time %in% c("1960")) %>% 
  group_by(name, time)
  ggplot(gdp_v_le_1960, aes(GDP_PC, lifeExpectancy, col=factor(region))) +
  geom_point(aes(size = totalPopulation/1000000)) +
  labs(title = "GDP per capita vs Life Expectancy by Country in 1960",
       x="GDP per capita (USD)",
       y="Life Expectancy (Years)",
       col = "Continents",
       size = "Total Population (Millions)")
```



2. (4pt) Make a similar plot, but this time use 2019 data only. \

```{r}
gdp_v_le_2019 <- population_data %>% 
  select(name, region, GDP_PC, lifeExpectancy, time, totalPopulation) %>% 
  filter(!is.na(name), !is.na(GDP_PC), !is.na(totalPopulation), !name=="", 
         time %in% c("2019")) %>% 
  group_by(name, time)
  ggplot(gdp_v_le_2019, aes(GDP_PC, lifeExpectancy, col=factor(region))) +
  geom_point(aes(size = totalPopulation/1000000)) +
  labs(title = "GDP per capita vs Life Expectancy by Country in 2019",
       x="GDP per capita (USD)",
       y="Life Expectancy (Years)",
       col = "Continents",
       size = "Total Population (Millions)")
```


3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years? \

The life expectancy and country populations have largely increased in general in the last 60 years,. However, the strongest outlier countries with very high GDP per capita have stayed around the same life expectancy even though they have increased in GDP per capita.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? \
Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\

```{r}
population_data %>% 
  select(region, lifeExpectancy, time) %>% 
  filter(!is.na(region), !region=="", time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  summarise(avg_life_expectancy = mean(lifeExpectancy, na.rm = TRUE)) 
```

  These results do confirm the results that we found in the graph. Continents with lower CO2 emissions like Africa and Asia in the 1960s experienced very significant life expectancy growth, while nations with already high life expectancy in the 1960s like the Americas and Europe experienced minimal life expectancy growth in 2016.


5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. \
Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\

```{r}
population_data %>% 
  select(region, lifeExpectancy, time) %>% 
  filter(!is.na(region), !region=="", time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  mutate(Growth = (lifeExpectancy - lag(lifeExpectancy))*100/lag(lifeExpectancy)) %>% 
  summarise(avg_growth = mean(Growth, na.rm = TRUE))
```

  This data shows that growth rates in regions with lower CO2 levels did experience faster life expectancy growth rates. However, in the 1960s it also states that the Americas also experienced a significant life expectancy growth. This can be interpreted as the large industrialization of Canada and the growth of Mexico. Now, all continents with the exception of Africa are experiencing stagnating life expectancies, a sign that perhaps the maximum life span of humanity has been reached with current science.

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it! \

```{r}
combo <- rbind(gdp_v_le_1960, gdp_v_le_2019)
ggplot(combo, aes(GDP_PC, lifeExpectancy, col=factor(time))) +
  geom_point() +
  labs(title = "GDP per capita vs Life Expectancy by Country in 1960 and 2016",
       x="GDP per capita (USD)",
       y="Life Expectancy (Years)",
       col = "Year")
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) \
Hint: check out the function rank()! \
Hint2: 17 for 1960. \

```{r}
population_data %>% 
  select(name, lifeExpectancy, time) %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), !name=="", time %in% c("1960", "2019")) %>% 
  arrange(time, desc(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(rankings = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America")
```


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? \
Hint: 0.0904 for 1960. \

```{r}
population_data %>% 
  select(name, lifeExpectancy, time) %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), !name=="", time %in% c("1960", "2019")) %>% 
  arrange(time, desc(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(rankings = rank(desc(lifeExpectancy))) %>% 
  select(name, time, rankings) %>% 
  group_by(time) %>% 
  mutate(total = max(rankings)) %>% 
  filter(name == "United States of America") %>% 
  summarise(relative_rank = rankings/total)
```



---------------------------------------------------------------------

Finally tell us how many hours did you spend on this PS.

This assignment was a lot more work than the previous Problem Sets. This PS5 probably took me close to around 8 hours throughout the week.

GITHUB REPO: https://github.com/DLJ949/ps5-gapminder/tree/main 
